Rotation Period Evolution in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2019-07-01 Paper/arxiv/sync.tex
added arxiv version of the paper

We plot the tidal evolution for $a$, $e$, and $P_{rot}$, ignoring stellar evolution, for a solar-twin binary with an initial P$_{orb} = 10$ d, P$_{rot} = 1$ d, $e = 0.2$ for the CPL model and CTL model, assuming $Q=10^6$ and $\tau = 0.1$ seconds, respectively, in Fig.~\ref{fig:tidalExample}. The CPL and CTL model predict the same qualitative evolution: both the binary's $e$ and P$_{orb}$ slightly increase as tides force the spins toward the tidally locked state, transferring rotational angular momentum into the orbit\xxx{, increasing the orbital angular momentum by ${\sim}1\%$} in the process.  At late times, both the CPL and CTL drive the binaries towards orbital circularization, with tidal dissipation decreasing P$_{orb}$. The predictions of the CPL and CTL model, differ, however, when the binaries tidally lock.  Under the CPL model, the binary tidally locks into a synchronous orbit when $e < \sqrt{1/19}$, e.g. Eq.~(\ref{eqn:cpl:eqPer}), while the CTL model predicts supersyncronous rotation due to the CTL model's equilibrium period eccentricity dependence, e.g. Eq.~(\ref{eqn:ctl:eqPer}).

@@

Our calculations show that subsynchronous rotation occurs across a wide range of tidal parameters and P$_{orb}$. In general as tides weaken, i.e. increasing P$_{orb}$ and/or \xxx{decreasing $\log_{10}(k_2 \tau)$}, tidal and magnetic braking torques balance at longer P$_{rot}$. \xxx{For strong tides, $\log_{10}(k_2 \tau) \gsim -1$}, tidal torques overpower magnetic braking for P$_{orb} \lsim 40$ d, tidally locking binaries into synchronous rotation. For our fiducial values of $k_2 = 0.5$ and $\tau = 0.1$ s (white dashed line in Fig.~\ref{fig:analyticBalance}), solar-twin binaries will rotate subsynchronously for P$_{orb} \gsim 20$ d, with more severe subsynchronism at longer P$_{orb}$. This simple calculation, however, does not account for stellar evolution or secular tidal orbital evolution, e.g. tidal friction that will shrink the orbit, gradually strengthening tidal torques, so we turn to simulations to characterize this evolution.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{eqPerTwoPanelMatt.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits according to the CPL (solid) and CTL (dashed) models with $Q = 10^6$ and $\tau = 0.1$ s, respectively, using the \citet{Matt2015} magnetic braking model. Left: P$_{rot}/$P$_{eq}$ for stars with P$_{orb}$ ranging from 5 d to 60 d. The black dotted line indicates the tidally locked state.  Right: Net P$_{rot}$ derivative due to stellar evolution, tidal torques, and magnetic braking.  We truncate each curve when the binary tidally locks. \xxx{The legend denotes the initial binary orbital period and we note that the orbital periods do not vary by more than a few percent over the course of the simulations.} For P$_{orb} \leq 10$ d, both the CPL and CTL models predict that the binaries lock into synchronous rotation.  For all P$_{orb}$, the CPL models tidally lock whereas the CTL model predicts subsynchronous rotation that persists for Gyrs.}%
    \label{fig:eqPer}%
\end{figure*}

@@

Spin-orbit synchronization is the typical outcome for binaries with P$_{orb} < 10$ d according to the CPL model for most values of $Q$. The strong tidal torques predicted by the CPL model can even tidally lock binaries out to P$_{orb} \gsim 80 $ d for $Q < 10^5$, well beyond the expected limit of 20 d \citep{Meibom2006}.  According to the CTL model, binaries with P$_{orb} < 10$ d typically tidally lock for $\tau \gsim 0.1$ s, and seldomly tidally lock for P$_{orb} > 20$ d, except for systems with strong tides, $\tau \gsim 3$ s.  Both models predict a substantial population of subsynchronous rotators (red regions in Fig.~\ref{fig:qTauLock}, P$_{rot} >$ P$_{orb}$), consistent with magnetic braking dominating weak tidal torques.  The population of supersynchronous rotators (blue regions in Fig.~\ref{fig:qTauLock}, P$_{rot} <$ P$_{orb}$) \xxx{with P$_{orb} > 60$ d} does not in general correspond to binaries tidally locking into supersynchronous rotation, but rather, typically arises from the combination of weak tidal torques and magnetic braking not spinning down stars enough for P$_{rot}$ to be close to the tidally locked state. \xxx{At a given age, longer P$_{orb}$ binaries will tend to rotate faster as they experience weaker tidal torques, and hence require longer to spin down towards the tidally locked state.}

@@

We isolate the impact of $Q$ and $\tau$ \xxx{on the spin-orbital state of tidally interacting stellar binaries} by binning our \xxx{CPL and CTL} simulation results after 7 Gyr of evolution by P$_{orb}$ and P$_{orb}/$P$_{rot}$ \xxx{in Figures \ref{fig:qmap} and \ref{fig:taumap}, respectively. In these figures, we estimate the typical strength of tidal torques, using $Q$ and $\tau$ as a proxy, that can produce various spin-orbital states. For the CPL simulations depicted in Fig.~\ref{fig:qmap},} synchronous and supersynchronous rotators have systematically low $Q$s, typically $Q < 10^6$, as strong tidal torques are required to tidally lock these binaries. For P$_{orb} < 10$ d, there are no rotators with $1.0 <$ P$_{orb}/$P$_{rot} < 1.5$, nor do any stars have P$_{orb}/$P$_{rot} > 1.5$ for P$_{orb} < 60$ d, as in the CPL model, binaries with eccentric orbits can only tidally lock into a 1:1 or 3:2 spin-orbit commensurability, see Eqn.~(\ref{eqn:cpl:eqPer}).

Subsynchronous rotators have systematically larger $Q$s, typically $Q > 10^6$, and hence experience weak tidal torques that are dominated by magnetic braking.  \xxx{Subsynchronous rotation can occur under the CPL model for binaries with P$_{orb} < 50$ d. In this regime, the median $Q$ tends to increase with decreasing P$_{orb}/$P$_{rot}$, except near the tidally locked state, as magnetic braking dominates weaker tidal torques, yielding longer P$_{rot}$.}  This trend reverses at longer P$_{orb} > 60$ d where supersynchronous rotation arises from the inability of tidal torques and magnetic braking to spin-down stars enough to approach the tidally locked state by the end of the simulation.  In this case, the more supersynchronous the rotation, the weaker the tidal torques must be, and hence the larger the $Q$ must be.

@@

According to the CTL model \xxx{simulations, depicted in Fig.~\ref{fig:taumap}}, many binaries tidally lock for P$_{orb} \lsim 20$ d when $\tau \gsim 0.1$ s, with some tidally locking up to P$_{orb} \approx 50$ d when $\tau \gsim 1$ s. Subsynchronous rotation typically occurs for stars with $\tau < 0.1$ s. \xxx{Similar to the behavior depicted in Fig.~\ref{fig:qmap}, longer P$_{rot}$ are produced by binaries with weaker tidal interactions since P$_{orb}/$P$_{rot}$ decreases monotonically with $\tau$ for P$_{orb} < 40$ d.}  For P$_{orb} > 50$ d, magnetic braking dominates the evolution \xxx{seen in} the diagonal sequence with a median $\tau \approx 0.1$ s, a value that is typically insufficient for tides to strongly influence the evolution given the wide orbital separations. The shape of this diagonal region arises from the combination of magnetic braking and our flat initial P$_{orb}$ distribution. \xxx{In this P$_{orb}$ regime, most binaries rotate supersynchronously as tides and magnetic braking fail to sufficiently spin down the stars by the age of the system. At longer P$_{orb}$, some binaries can strongly tidally-interact, but these systems require extreme tidal $\tau \gsim 10$ s. Unlike the CPL simulations depicted in Fig.~\ref{fig:qmap}, these binaries with P$_{orb} \approx 90$ d are not tidally locked as the tides are still not strong enough to lock the system. We explore this point further in $\S$~\ref{sec:protDist}.} We do not often observe P$_{orb}/$P$_{rot} \gsim 1.5$ as we only consider eccentricities up to $e = 0.3$, limiting how rapid supersynchronous systems can rotate according to Eqn.~(\ref{eqn:ctl:eqPer}).

@@

The interacting population tends to rotate more slowly than the not locked population as at short P$_{orb}$, and hence P$_{rot}$, binaries preferentially tidally lock due to stronger tidal torques.  At longer P$_{orb}$, weaker tidal torques allow magnetic braking to spin down the stars past P$_{eq}$, with tidal torques eventually strengthening enough to shepherd P$_{rot}$ towards P$_{eq}$ via the mechanism discussed in $\S$~\ref{sec:eq}. The CPL and CTL models predict that \xxx{$31\%$ and $24\%$} of stars, respectively, are either tidally locked or interacting, demonstrating that tidal torques play a pivotal role in shaping the angular momentum evolution in stellar binaries across a wide range of parameters. The P$_{rot}$ - mass distribution for not locked binaries resembles the single star sequence as magnetic braking and stellar evolution dictate their angular momentum evolution.

Tidal locking is not limited to P$_{orb} \lsim 20$ d, however, as we find stellar binaries can tidally lock over a wide range of P$_{rot}$ up to P$_{rot} =$ P$_{orb} \approx 100$ d according to the CPL model, producing a slow-rotating population above the P$_{rot}$ distribution envelop of solar-mass single stars. This behavior is consistent with observations of P$_{rot}$ in \kepler eclipsing binaries by \citet{Lurie2017} who find tentative evidence that binaries can tidally lock up to their detection limit of P$_{orb} = $ P$_{rot} = 45$ d. Under the CTL model, however, binaries predominantly tidally lock out to only P$_{orb} \approx 20$ d, although binaries with more massive primaries can occasionally lock, or at least tidally-interact, out to P$_{orb} \approx 80$ d. We highlight this behavior with a histogram of locked and interacting binaries over P$_{rot}$ for both tidal models in Fig.~\ref{fig:lockedProtHist}. \xxx{The CTL model predicts fewer tidally locked binaries at longer P$_{rot}$, concentrating most of the locked CTL distribution's density at short P$_{rot}$, whereas the CPL distribution has a heavy tail extending towards longer P$_{rot}$. The CPL model, however, predicts larger tidal-locking rates than the CTL model as seen in the enhanced numbers of tidally locked binaries at low P$_{rot}$ Fig.~\ref{fig:lockedCPL} compared with Fig.~\ref{fig:lockedCTL}.} The presence of P$_{orb} > 20$ d locked population, or lack there of, could be a powerful observational discriminant between which equilibrium tidal model acts in low-mass stellar binaries. We discuss this point further in $\S$~\ref{sec:whichModel}.

>>>>

We plot the tidal evolution for $a$, $e$, and $P_{rot}$, ignoring stellar evolution, for a solar-twin binary with an initial P$_{orb} = 10$ d, P$_{rot} = 1$ d, $e = 0.2$ for the CPL model and CTL model, assuming $Q=10^6$ and $\tau = 0.1$ seconds, respectively, in Fig.~\ref{fig:tidalExample}. The CPL and CTL model predict the same qualitative evolution: both the binary's $e$ and P$_{orb}$ slightly increase as tides force the spins toward the tidally locked state, transferring rotational angular momentum into the orbit, increasing the orbital angular momentum by ${\sim}1\%$ in the process.  At late times, both the CPL and CTL drive the binaries towards orbital circularization, with tidal dissipation decreasing P$_{orb}$. The predictions of the CPL and CTL model, differ, however, when the binaries tidally lock.  Under the CPL model, the binary tidally locks into a synchronous orbit when $e < \sqrt{1/19}$, e.g. Eq.~(\ref{eqn:cpl:eqPer}), while the CTL model predicts supersyncronous rotation due to the CTL model's equilibrium period eccentricity dependence, e.g. Eq.~(\ref{eqn:ctl:eqPer}).

@@

Our calculations show that subsynchronous rotation occurs across a wide range of tidal parameters and P$_{orb}$. In general as tides weaken, i.e. increasing P$_{orb}$ and/or decreasing $\log_{10}(k_2 \tau)$, tidal and magnetic braking torques balance at longer P$_{rot}$. For strong tides, $\log_{10}(k_2 \tau) \gsim -1$, tidal torques overpower magnetic braking for P$_{orb} \lsim 40$ d, tidally locking binaries into synchronous rotation. For our fiducial values of $k_2 = 0.5$ and $\tau = 0.1$ s (white dashed line in Fig.~\ref{fig:analyticBalance}), solar-twin binaries will rotate subsynchronously for P$_{orb} \gsim 20$ d, with more severe subsynchronism at longer P$_{orb}$. This simple calculation, however, does not account for stellar evolution or secular tidal orbital evolution, e.g. tidal friction that will shrink the orbit, gradually strengthening tidal torques, so we turn to simulations to characterize this evolution.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{eqPerTwoPanelMatt.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits according to the CPL (solid) and CTL (dashed) models with $Q = 10^6$ and $\tau = 0.1$ s, respectively, using the \citet{Matt2015} magnetic braking model. Left: P$_{rot}/$P$_{eq}$ for stars with P$_{orb}$ ranging from 5 d to 60 d. The black dotted line indicates the tidally locked state.  Right: Net P$_{rot}$ derivative due to stellar evolution, tidal torques, and magnetic braking.  We truncate each curve when the binary tidally locks. The legend denotes the initial binary orbital period and we note that the orbital periods do not vary by more than a few percent over the course of the simulations.} For P$_{orb} \leq 10$ d, both the CPL and CTL models predict that the binaries lock into synchronous rotation.  For all P$_{orb}$, the CPL models tidally lock whereas the CTL model predicts subsynchronous rotation that persists for Gyrs.%
    \label{fig:eqPer}%
\end{figure*}

@@

Spin-orbit synchronization is the typical outcome for binaries with P$_{orb} < 10$ d according to the CPL model for most values of $Q$. The strong tidal torques predicted by the CPL model can even tidally lock binaries out to P$_{orb} \gsim 80 $ d for $Q < 10^5$, well beyond the expected limit of 20 d \citep{Meibom2006}.  According to the CTL model, binaries with P$_{orb} < 10$ d typically tidally lock for $\tau \gsim 0.1$ s, and seldomly tidally lock for P$_{orb} > 20$ d, except for systems with strong tides, $\tau \gsim 3$ s.  Both models predict a substantial population of subsynchronous rotators (red regions in Fig.~\ref{fig:qTauLock}, P$_{rot} >$ P$_{orb}$), consistent with magnetic braking dominating weak tidal torques.  The population of supersynchronous rotators (blue regions in Fig.~\ref{fig:qTauLock}, P$_{rot} <$ P$_{orb}$) with P$_{orb} > 60$ d does not in general correspond to binaries tidally locking into supersynchronous rotation, but rather, typically arises from the combination of weak tidal torques and magnetic braking not spinning down stars enough for P$_{rot}$ to be close to the tidally locked state. At a given age, longer P$_{orb}$ binaries will tend to rotate faster as they experience weaker tidal torques, and hence require longer to spin down towards the tidally locked state.

@@

We isolate the impact of $Q$ and $\tau$ on the spin-orbital state of tidally interacting stellar binaries by binning our CPL and CTL simulation results after 7 Gyr of evolution by P$_{orb}$ and P$_{orb}/$P$_{rot}$ in Figures \ref{fig:qmap} and \ref{fig:taumap}, respectively. In these figures, we estimate the typical strength of tidal torques, using $Q$ and $\tau$ as a proxy, that can produce various spin-orbital states. For the CPL simulations depicted in Fig.~\ref{fig:qmap}, synchronous and supersynchronous rotators have systematically low $Q$s, typically $Q < 10^6$, as strong tidal torques are required to tidally lock these binaries. For P$_{orb} < 10$ d, there are no rotators with $1.0 <$ P$_{orb}/$P$_{rot} < 1.5$, nor do any stars have P$_{orb}/$P$_{rot} > 1.5$ for P$_{orb} < 60$ d, as in the CPL model, binaries with eccentric orbits can only tidally lock into a 1:1 or 3:2 spin-orbit commensurability, see Eqn.~(\ref{eqn:cpl:eqPer}).

Subsynchronous rotators have systematically larger $Q$s, typically $Q > 10^6$, and hence experience weak tidal torques that are dominated by magnetic braking.  Subsynchronous rotation can occur under the CPL model for binaries with P$_{orb} < 50$ d. In this regime, the median $Q$ tends to increase with decreasing P$_{orb}/$P$_{rot}$, except near the tidally locked state, as magnetic braking dominates weaker tidal torques, yielding longer P$_{rot}$.  This trend reverses at longer P$_{orb} > 60$ d where supersynchronous rotation arises from the inability of tidal torques and magnetic braking to spin-down stars enough to approach the tidally locked state by the end of the simulation.  In this case, the more supersynchronous the rotation, the weaker the tidal torques must be, and hence the larger the $Q$ must be.

@@

According to the CTL model simulations, depicted in Fig.~\ref{fig:taumap}, many binaries tidally lock for P$_{orb} \lsim 20$ d when $\tau \gsim 0.1$ s, with some tidally locking up to P$_{orb} \approx 50$ d when $\tau \gsim 1$ s. Subsynchronous rotation typically occurs for stars with $\tau < 0.1$ s. Similar to the behavior depicted in Fig.~\ref{fig:qmap}, longer P$_{rot}$ are produced by binaries with weaker tidal interactions since P$_{orb}/$P$_{rot}$ decreases monotonically with $\tau$ for P$_{orb} < 40$ d.  For P$_{orb} > 50$ d, magnetic braking dominates the evolution seen in the diagonal sequence with a median $\tau \approx 0.1$ s, a value that is typically insufficient for tides to strongly influence the evolution given the wide orbital separations. The shape of this diagonal region arises from the combination of magnetic braking and our flat initial P$_{orb}$ distribution. In this P$_{orb}$ regime, most binaries rotate supersynchronously as tides and magnetic braking fail to sufficiently spin down the stars by the age of the system. At longer P$_{orb}$, some binaries can strongly tidally-interact, but these systems require extreme tidal $\tau \gsim 10$ s. Unlike the CPL simulations depicted in Fig.~\ref{fig:qmap}, these binaries with P$_{orb} \approx 90$ d are not tidally locked as the tides are still not strong enough to lock the system. We explore this point further in $\S$~\ref{sec:protDist}. We do not often observe P$_{orb}/$P$_{rot} \gsim 1.5$ as we only consider eccentricities up to $e = 0.3$, limiting how rapid supersynchronous systems can rotate according to Eqn.~(\ref{eqn:ctl:eqPer}).

@@

The interacting population tends to rotate more slowly than the not locked population as at short P$_{orb}$, and hence P$_{rot}$, binaries preferentially tidally lock due to stronger tidal torques.  At longer P$_{orb}$, weaker tidal torques allow magnetic braking to spin down the stars past P$_{eq}$, with tidal torques eventually strengthening enough to shepherd P$_{rot}$ towards P$_{eq}$ via the mechanism discussed in $\S$~\ref{sec:eq}. The CPL and CTL models predict that $31\%$ and $24\%$ of stars, respectively, are either tidally locked or interacting, demonstrating that tidal torques play a pivotal role in shaping the angular momentum evolution in stellar binaries across a wide range of parameters. The P$_{rot}$ - mass distribution for not locked binaries resembles the single star sequence as magnetic braking and stellar evolution dictate their angular momentum evolution.

Tidal locking is not limited to P$_{orb} \lsim 20$ d, however, as we find stellar binaries can tidally lock over a wide range of P$_{rot}$ up to P$_{rot} =$ P$_{orb} \approx 100$ d according to the CPL model, producing a slow-rotating population above the P$_{rot}$ distribution envelop of solar-mass single stars. This behavior is consistent with observations of P$_{rot}$ in \kepler eclipsing binaries by \citet{Lurie2017} who find tentative evidence that binaries can tidally lock up to their detection limit of P$_{orb} = $ P$_{rot} = 45$ d. Under the CTL model, however, binaries predominantly tidally lock out to only P$_{orb} \approx 20$ d, although binaries with more massive primaries can occasionally lock, or at least tidally-interact, out to P$_{orb} \approx 80$ d. We highlight this behavior with a histogram of locked and interacting binaries over P$_{rot}$ for both tidal models in Fig.~\ref{fig:lockedProtHist}. The CTL model predicts fewer tidally locked binaries at longer P$_{rot}$, concentrating most of the locked CTL distribution's density at short P$_{rot}$, whereas the CPL distribution has a heavy tail extending towards longer P$_{rot}$. The CPL model, however, predicts larger tidal-locking rates than the CTL model as seen in the enhanced numbers of tidally locked binaries at low P$_{rot}$ Fig.~\ref{fig:lockedCPL} compared with Fig.~\ref{fig:lockedCTL}. The presence of P$_{orb} > 20$ d locked population, or lack there of, could be a powerful observational discriminant between which equilibrium tidal model acts in low-mass stellar binaries. We discuss this point further in $\S$~\ref{sec:whichModel}.
